Iterative Monomorphisation

### 2024-03-20 iterative_monomorphisation.tex
edits

\setlist[itemize]{topsep=3pt, itemsep=3pt}
\setlist[enumerate]{topsep=3pt, itemsep=3pt}

@@

The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a lower level algorithm with the following features.
\begin{enumerate}
    \item[\labelitemi] Bounds are presented at several stages to curb potentially explosive enumerations.

    \item[\labelitemi] Type arguments are separated into old and new sets to avoid re-computing some of the same matchings in successive iterations.

    \item[\labelitemi] Substitutions are directly applied to the type arguments instead of the formulae. This dispenses from having to re-extract the type arguments from the formulae at each iteration. New formulae are only generated once all iterations are complete, in a separate, final step.

The data structures used in the pseudocode are based on the ones used in the high-level description. Instead of \(M\) a map from symbols to monomorphic type argument tuples, we have \(M_\text{old}\) and \(M_\text{new}\) which play the same role whilst also distinguishing between type arguments having already been matched against and those that have not. Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace \(N\) the map from formulae to symbols to non-monomorphic type argument tuples. Finally, we keep track of a set \(S\) of substitutions generated by the matchings, until it is used to generate new formulae in the final phase.

All sets referenced in the pseudocode are assumed to be finite. Additionally, this algorithm relies on ``primitives'' whose implementation depend on the specifics of the grammar and logic used. They will therefore not be expanded upon. Functions computing the following are assumed to be available: 
\begin{enumerate}
   \item[\labelitemi] \emph{initialisation\((\Phi)\)} with \(\Phi\) a set of (polymorphic) formulae extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
   \item[\labelitemi] \emph{type\_variables\(((\tau_1, \dots, \tau_n))\)} with \((\tau_1, \dots,\tau_n)\) a type tuple gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set.
   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)} with \(\upsilon\) and \(\tau\) types will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)} where \(\sigma\) is the substitution resulting from the matching. Note that the pseudocode only matches non-monomorphic types against monomorphic types, in practice it is sufficient for \emph{match} to only cover this case.
   \item[\labelitemi] \emph{domain\((\sigma)\)} where \(\sigma\) is a substitution returns the set of type variables instantiated by \(\sigma\).
   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} where \(\sigma_1\) and \(\sigma_2\) are substitutions tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
   \item[\labelitemi] Substitution composition and application are noted mathematically but still need to be implemented.
\end{enumerate}

@@

The iterative monomorphisation algorithm (\ref{iter_mono}), as implemented in Zipperposition, has two main phases. The first phase applies a \emph{monomorphisation step} (\ref{mono_step}) to each formula in \(\Phi\) until the user-set limit it reached. This limit is the only bound necessary for the algorithm to terminate. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once the first phase is completed, the substitutions used to create new type arguments are passed to the formula generation function (\ref{gen_formulae}) for the second phase.

@@

Formula monomorphisation (\ref{mono_step}) computes new type arguments for a single formula, it is the core of the monomorphisation process. The first step generates substitutions (\ref{subst_gen}) by matching the non-monomorphic type arguments against monomorphic ones. The separation of type arguments into ``old'' and ``new'' maps is put to use here. Pairs of type argument tuples must contain at least one new type argument in order to be matched. This avoids re-computing some of the matchings processed in previous iterations. Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.

The set of substitutions generated thusly is then iterated over and each substitution is applied to all polymorphic type arguments. The application of a substitution to a tuple of polymorphic type arguments only takes place if least one type variable in the tuple is instantiated by the substitution. This condition filters out all substitution application that would have otherwise left all type arguments unchanged.

@@

Substitution generation (\ref{subst_gen}) consists in computing all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions. For each pair of type argument tuples, the non-monomorphic types are matched against their monomorphic counterparts and the resulting substitutions are composed if they are compatible. In the pseudocode, this is checked by making sure that the while loop has iterated over all elements of the type argument tuple. If any substitutions are incompatible, the matchings are discarded. The composition of two compatible substitutions is commutative and that the order of composition is therefore irrelevant.

@@

To monomorphise a polymorphic formula, we first compute its \emph{monomorphising substitutions} (\ref{mono_substs}). A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic. Such substitutions are computing using a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) which instantiates at least one of the type variables in \(V\). It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they may be composed and the function recursively called to instantiate the remaining type variables.

@@

Finally, once all monomorphising substitutions have been generated, they are applied to the polymorphic formula of the input problem that they instantiate (\ref{gen_formulae}). The fact that the substitutions are monomorphising relative to the formula they applied to ensures that the resulting formulae are monomorphic. A final bound here is used to control the total number of new formulae. It overlaps in role with the bound limiting the number of monomorphising substitutions per clause, but can be useful as a means to set an absolute cap on the size of the final problem. In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of clauses it is given.

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, several competitive higher-order polymorphic provers were run on the set of TPTP test problems. The total number of problems solved the ``Native'' results shown in the table. Evaluation of the monomorphisation approach is done in two steps, first each problem is monomorphised using the [insert configuration] configurations [options?]. Secondly, each prover is ran on the monomorphised problem and the results are tallied in the ``Mono'' column of the table. In addition to the polymorphic provers used in the ``Native'' tests, two monomorphic provers are run on the monomorphised problems. Note that instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time is subtracted in order to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.

>>>>

\setlist[itemize]{topsep=5pt, itemsep=0pt}
\setlist[enumerate]{topsep=5pt, itemsep=0pt}

@@

The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a lower level algorithm with the following features.
First, numeric bounds are introduced to stop explosive enumerations.
Second, type arguments are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
Third, substitutions are directly applied to the type arguments instead of the formulae. This dispenses from having to re-extract the type arguments from the formulae at each iteration. New formulae are generated only once all iterations are complete, in a separate, final step.

The data structures used in the pseudocode are based on the ones used in the high level description. Instead of a map \(M\) from symbols to monomorphic type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type arguments having already been matched against and those that have not. Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples. Finally, we keep track of a set \(S\) of substitutions generated by the matchings. It is used to generate new formulae in the final phase.

All sets referenced in the pseudocode are assumed to be finite. Additionally, this algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used. They will therefore not be expanded upon. Functions computing the following are assumed to be available: 
\begin{enumerate}
   \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set.
   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)}, where \(\sigma\) is the substitution resulting from the matching. The pseudocode matches only non-monomorphic types against monomorphic types.
   \item[\labelitemi] \emph{domain\((\sigma)\)}, where \(\sigma\) is a substitution, returns the set of type variables \(\alpha\) such that \(\sigma(
   \alpha) \not= \alpha\).
   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)}, where \(\sigma_1\) and \(\sigma_2\) are substitutions, tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
\end{enumerate}

@@

The iterative monomorphisation algorithm (Figure~\ref{iter_mono}) has two phases. The first phase applies a \emph{monomorphisation step} (Figure~\ref{mono_step}) to each formula in \(\Phi\) until the user-set limit is reached. This limit is the only bound necessary for the algorithm to terminate. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is completed and the substitutions used to create
new type arguments are passed to the formula generation function
(Figure~\ref{gen_formulae}) for the second phase.

@@

Formula monomorphisation (Figure~\ref{mono_step}) computes new type arguments for a single formula. It forms the core of the monomorphisation process. The first step generates substitutions (Figure~\ref{subst_gen}) by matching the non-monomorphic type arguments against monomorphic ones. The separation of type arguments into old and new maps is put to use here. Pairs of type argument tuples must contain at least one new type argument to be matched. This avoids re-computing some of the matchings processed in previous iterations. Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.

The set of substitutions generated in this way is then iterated over, and each substitution is tentatively applied to all polymorphic type arguments. The application of a substitution to a tuple of polymorphic type arguments takes place only if least one type variable in the tuple is instantiated by the substitution. This condition filters out all substitution application that would have otherwise left all type arguments unchanged.

@@

Substitution generation (Figure~\ref{subst_gen}) consists in computing all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions. For each pair of type argument tuples, the non-monomorphic types are matched against their monomorphic counterparts and the resulting substitutions are composed if they are compatible. In the pseudocode, this is checked by making sure that the while loop has iterated over all elements of the type argument tuple. If any substitutions are incompatible, the matchings are discarded. The composition of two compatible substitutions is commutative and that the order of composition is therefore irrelevant.

@@

To monomorphise a polymorphic formula, we first compute its \emph{monomorphising substitutions} (Figure~\ref{mono_substs}). A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic. Such substitutions are computing using a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) which instantiates at least one of the type variables in \(V\). It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they may be composed and the function recursively called to instantiate the remaining type variables.

@@

Finally, once all monomorphising substitutions have been generated, they are applied to the polymorphic formula of the input problem that they instantiate (Figure~\ref{gen_formulae}). The fact that the substitutions are monomorphising relative to the formula they applied to ensures that the resulting formulae are monomorphic. A final bound here is used to control the total number of new formulae. It overlaps in role with the bound limiting the number of monomorphising substitutions per clause, but can be useful as a means to set an absolute cap on the size of the final problem. In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of clauses it is given.

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, several competitive higher-order polymorphic provers were run on the set of TPTP test problems. The total number of problems solved the ``Native'' results shown in the table. Evaluation of the monomorphisation approach is done in two steps, first each problem is monomorphised using the [insert configuration] configurations [options?]. Secondly, each prover is ran on the monomorphised problem and the results are tallied in the ``Mono'' column of the table. In addition to the polymorphic provers used in the ``Native'' tests, two monomorphic provers are run on the monomorphised problems. Note that instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time is subtracted to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.
